On the Minimum Core Mass for Giant Planet Formation

### 2013-06-17 tex/draft/draft2.tex
current paper versions

\newcommand{\Eq}[1]{Equation\,(\ref{#1})}
\newcommand{\Eqs}[2]{Equations (\ref{#1}) and~(\ref{#2})}
\newcommand{\Eqss}[2]{Equations (\ref{#1})--(\ref{#2})}
\newcommand{\Eqsss}[3]{Equations (\ref{#1}), (\ref{#2}) and~(\ref{#3})}
\newcommand{\App}[1]{Appendix~\ref{#1}}
\newcommand{\Sec}[1]{Sect.~\ref{#1}}
\newcommand{\Chap}[1]{Chapter~\ref{#1}}
\newcommand{\Fig}[1]{Fig.~\ref{#1}}
\newcommand{\Figs}[2]{Figs.~\ref{#1} and \ref{#2}}
\newcommand{\Figss}[2]{Figs.~\ref{#1}--\ref{#2}} 
\newcommand{\Tab}[1]{Table \ref{#1}}

@@

Standard core accretion models (e.g., \citealt{stevenson82}, \citealt{rafikov06}) assume that the core and the atmosphere grow simultaneously, and that planetesimal accretion deposits enough heat to alter the luminosity of the atmosphere, increasing the core mass required for the atmosphere to collapse. These studies consider that the planet atmosphere is in a steady state at all times, where all the incoming energy from planetesimal accretion is radiated away by the envelope, and  find that there exists a minimum (``critical'') core mass past which hydrostatic solutions can no longer be found and unstable atmosphere collapse occurs.  Standard calculations find the critical core mass to be about $10 M_{\oplus}$ (e.g., \citealt{stevenson82}, \citealt{rafikov06}).

@@

The time dependence of the atmosphere structure equations may be neglected or explicitly taken into account. Some previous studies of atmosphere accretion (e.g., \citealt{stevenson82}, \citealt{wuchterl93}, \citealt{rafikov06}) consider static envelopes, in which the luminosity is solely supplied by planetesimal accretion and fully radiated away by the atmosphere. In other studies, the time evolution is explicitly taken into account and full time dependent models are developed (e.g., \citealt{ikoma00}). We follow an intermediate approach and consider quasi-static evolution. Our model for the atmosphere growth time is described in section \ref{cooling}.

@@

\begin{subeqnarray}
\label{eq:struct}
\frac{dP}{dr}&=&-\frac{G m}{r^2}\rho \slabel{eq:structa} \\
\frac{dm}{dr}&=&4 \pi r^2 \rho\slabel{eq:structb} \\
\frac{dT}{dr}&=&\nabla \frac{T}{P}\frac{dP}{dr}\slabel{eq:structc} \\
\frac{dL}{dr}&=&4 \pi r^2 \rho \epsilon_g\slabel{eq:structd}, 
\end{subeqnarray}

\noindent where $r$ is the radial coordinate, $P$, $T$ and $\rho$ are the gas pressure, temperature and density, respectively, $m$ is the mass enclosed by the radius $r$, $L$ is the luminosity from the surface of radius $r$, and $\epsilon_g \equiv -T \frac{ds}{dt}$ represents the heating per unit mass due to gravitational contraction, with $s$ the specific gas entropy. The temperature gradient $\nabla \equiv \frac{d \ln T}{d \ln P}$ has different expressions depending on the primary means of energy transport throughout the atmosphere. We assume that energy can be transported either through radiation or convection. When the luminosity is carried by radiative diffusion, the temperature gradient is given by

@@

\noindent where $\sigma$ is the Stefan-Boltzmann constant and $\kappa$ is the opacity. Alternatively, when the energy is transported outwards through convective motions, the temperature gradient becomes

@@

\noindent where  $\mathcal{R}$ is the reduced gas constant: $\mathcal{R}=k_b/(\mu m_p)$, with $k_b$ the Boltzmann constant and $m_p$ the proton mass. The gas constant $\mathcal{R}$ depends on the mean molecular weight of the gas $\mu$, which is determined by the relative abundance of hydrogen and helium in the atmosphere. In this study we assume that the helium fraction only affects the molecular weight of the gas, and not the adiabatic index. We show in section \ref{critcore} that this assumption does not affect the qualitative behavior of our solutions.

@@

\begin{equation}
\label{eq:RHill}
R_{\rm H} \equiv a \Big(\frac{M_{\rm p}}{3 M_{\odot}}\Big)^{1/3}
\end{equation}

\noindent Outside the Hill radius, the gravity of the planet is too weak to affect the nebular gas. Consequently, the planet will be able to accrete only gas that lies within its Hill sphere. \\

@@

\begin{subeqnarray}
\label{eq:diskparam}
\Sigma\di&=&2200 F_{\Sigma} a^{-3/2}\,\, \text{g cm}^{-2} \slabel{eq:diska}\\
T\di &=& 120 F_T a^{-3/7} \,\,K, \slabel{eq:diskb}
\end{subeqnarray}

\noindent with $a$ the semi-major axis in AU, and $F_{\Sigma}$ and $F_T$ normalization factors that adjust the disk mass and temperature relative to the minimum mass solar nebula (MMSN). In this study, we assume $F_{\Sigma}=F_T=1$. The resulting mid-plane pressure and disk scale height are given by

\begin{equation}
\label{eq:Pd}
P\di=110  F_{\Sigma} \sqrt{F_T m_*} a^{-45/14} \,\, \text{dyn cm}^{-2}
\end{equation}
and
\begin{equation}
H=0.022 \sqrt{F_T} a^{9/7}\,\, \text{AU},
\end{equation}

@@

\noindent where the total luminosity $L$ is balanced by the rate of heat generation $\Gamma$ (due to nuclear fusion in the case of star) and the rate at which total energy is lost $\dot{E}$. The total energy of the sphere is given by the sum of the gravitational and internal energies.

@@

This cooling model applies at any radius $R$ where the mass enclosed is $M$, for example the Bondi radius or the Hill radius. $L$, $\Gamma$ and $\dot{E}$ are the same as in equation (\ref{eq:coolingstar}), and include the atmosphere from the core to the top boundary. The gravitational energy $E_G$ of an atmosphere with total mass $M$ is given by

\begin{equation}
\label{eq:Eg}
E_G=-\int_{M\co}^M \frac{G m}{r} dm,
\end{equation}

\noindent while the internal energy is

\begin{equation}
\label{eq:U}
U=\int_{M\co}^M u dm,
\end{equation}

\noindent where $u$ is the internal energy per mass. We now explain the additional terms in the cooling equation (\ref{eq:coolingglobal}). $L\co$ is the luminosity from the solid core, and may include planetesimal accretion and radioactive decay. $e_{\mathrm{acc}}$ is the specific total energy brought in by mass accreting at the rate $\dot{M}$: $e_{\mathrm{acc}}=u-G M/R$. The last term represents the work done on a mass element as the atmosphere contracts.

The global cooling equation (\ref{eq:coolingglobal}) can be derived starting from the local cooling balance. Integrating the local energy equation and (\ref{eq:structd}) from the core to the surface and using equation (\ref{eq:structb}) yields

\begin{subeqnarray}
\label{eq:coolinglocal}
L-L_c&=&\int_{M_c}^M \frac{\partial L}{\partial m} dm \\
&=&\int_{M_c}^M\Big(\epsilon - T \frac{\partial s}{\partial t}\Big) dm \slabel{eq:clb} \\
&=&\Gamma-\int_{M_c}^M\frac{\partial u}{\partial t}dm + \int_{M_c}^M \frac{P}{\rho^2}\frac{\partial{\rho}}{\partial t} dm \slabel{eq:clc},
\end{subeqnarray}

\noindent where $\Gamma=\int_{M_c}^M \epsilon dm$ and equation (\ref{eq:clc}) follows from equation (\ref{eq:clb}) through the first law of thermodynamics.

Moreover, a self-gravitating protoplanetary atmosphere has to satisfy virial equilibrium. The virial theorem is derived by integrating the equation of hydrostatic balance (\ref{eq:structa}) and has the form

\begin{equation}
\label{eq:virial}
E_G=-3 \int_{M_c}^M \frac{P}{\rho} dm + 4 \pi (R^3 P_M-R_c^3 Pc)
\end{equation}

Similarly to the cooling equation, it applies at any radius $R$ where the total mass enclosed is $M$. For an ideal gas of adiabatic index $\gamma$, equation (\ref{eq:virial}) leads to the following expression for the total energy:

\begin{subeqnarray}
E&=&(1-\xi)U+4 \pi (R^3 P_M-R_c^3 Pc) \slabel{eq:vira} \\
&=&\frac{\xi-1}{\xi}E_G+\frac{4 \pi}{\xi} (R^3 P_M-R_c^3 Pc) \slabel{eq:virb}
\end{subeqnarray}

The global cooling equation (\ref{eq:coolingglobal}) can therefore be derived using equations(\ref{eq:coolinglocal}) and (\ref{eq:virial}). A detailed derivation is deferred to  \App{sec:virial}.

@@

In the next subsections we derive the structure and cooling behavior of such an atmosphere, deferring some of the calculations to \App{sec:analytic}. We estimate the temperature and pressure corrections at the radiative-convective boundary due to the fact that the outer radiative region of the atmosphere is not purely isothermal in section \ref{RCBcorr}. Finally, we derive the time evolution of the atmosphere and determine the minimum core mass for runaway gas accretion in section \ref{coolingan}.

@@

We use the boundary conditions and opacity laws described in section \ref{struct}. The numerical integration is performed through the shooting method, which solves the boundary value problem by reducing it to an initial value problem: trial values are chosen for the parameters at one of the boundaries, then the equations are integrated and the resulting values at the other boundary are compared to the actual boundary conditions. The procedure is repeated until convergence is achieved. We start with a total atmosphere mass $M_i$, where the index $i$ labels each evolutionary stage. For this mass we guess a trial luminosity $L=L_{\mathrm{guess}}$. The assumption of constant luminosity throughout the radiative zone sets the right hand side of equation (\ref{eq:structb}) to zero, i.e. the time dependence is neglected. We integrate the structure equations (\ref{eq:struct}) inwards from $R_{\mathrm{out}}=R_{\rm H}$, with the boundary conditions $T(R_{\mathrm{out}})=T_{\rm d}$ and $P(R_{\mathrm{out}})=P_{\rm d}$. The numerical integration gives a value for the core mass implied by the trial solution $M_{\rm c}=M_{\rm{c, guess}}$. We adjust $L_{\mathrm{guess}}$ until $M_{\rm{c,guess}}$ converges to the actual core mass $M_{\rm c}$.

The time evolution of the atmosphere is obtained using the global cooling model described in section \ref{cooling}. We neglect the luminosity due to planetesimal accretion and direct heat generation; as such, the first two terms in equation (\ref{eq:coolingglobal}) are set to zero. Equation (\ref{eq:coolingglobal}) is then evaluated at the RCB, since we assume that luminosity is primarily generated in the convective zone. In what follows, $X_{RCB}$ denotes quantity $X$ evaluated at the RCB. The time step $\Delta t_i$ between two consecutive models of masses $M_i$ and $M_{i+1}$ is calculated as follows:

\begin{eqnarray}\nonumber
\label{eq:dti}
<L>_i \Delta t_i&=&-\Delta E_{RCB,i} + <e_{\mathrm{acc}, RCB}>_i \Delta M_{RCB, i} \\
&&  - <P_{RCB}>_i \Delta V_{RCB, M, i} 
\end{eqnarray}

@@

We consider planets embedded in a protoplanetary disk around a Sun-like star and with a temperature and pressure profile described by equations (\ref{eq:diskb}) and (\ref{eq:Pd}), respectively. We are primarily interested in the outer regions of the disk, so we generate atmosphere models at distances between 10 and 100 AU in our fiducial disk. We emphasize that the location in the disk only affects the temperature and pressure of the nebular gas. As our model applies only when planetesimal accretion occurs too slowly to substantially change the mass of the core, we consider the evolution of an atmosphere around a core of fixed mass.

@@

\begin{figure}[h]
\centering
\includegraphics[width=0.5\textwidth]{../../figs/ModelAtmospheres/RadSelfGravPoly/PaperFigs/PTm_profiles_v2.pdf}
%\vspace{-0.5in}
\caption{Example pressure, temperature and total mass (core plus atmosphere) profiles as a function of radius at $a=60$ AU and for a fixed core mass $M_{\rm c}=5 M_{\oplus}$. The circles and triangles mark the locations of the Bondi radius and of the radiative-convective boundary, respectively. The solid, dotted and dashed lines correspond to profiles with total mass (core and atmosphere) of $5.12 M_{\oplus}$, $5.85 M_{\oplus}$ and $9.22 M_{\oplus}$, respectively.}
\label{fig:profiles}
\end{figure}

@@

In what follows we discuss the time evolution of the envelope. We use the global cooling model developed in section \ref{cooling} to obtain an evolutionary history from series of static atmosphere profiles, using the numerical procedure described in section \ref{twolayer}. Figure \ref{fig:Ltplot} shows the luminosity evolution and the elapsed time as a function of atmosphere mass for the same choice of parameters as in Figure \ref{fig:profiles} ($M_{\rm c}=5 M_{\oplus}$). The radiative region of the atmosphere is initially thin, which results in a relatively small number of steps that photons need to diffuse and hence a high luminosity (see equation \ref{eq:Lcb}). The luminosity decreases as the atmosphere mass grows and the radiative region becomes deeper. For comparison, we also plot the analytic results derived in section \ref{analytic}. We find only a brief agreement between the analytic and numerical results. In the early stages, the radiative zone is too shallow and the assumption that $P_{\rm RCB} \gg P_{\rm d}$ used in the analytic model breaks down. As the mass of the atmosphere grows, the neglect of self-gravity in the analytic model results in a lower luminosity (equation \ref{eq:Lcb}) and therefore a longer cooling time when compared to the numerical solution. While our analytic model does not provide an accurate quantitative description of the atmosphere evolution, we will show in section \ref{critical} that it predicts the correct scalings for the atmosphere crossover time and critical core mass.  %We see that self-gravity starts to matter

\begin{figure}[h]
\centering
\includegraphics[width=0.5\textwidth]{../../figs/ModelAtmospheres/RadSelfGravPoly/PaperFigs/Lt_profiles_v2.pdf}
%\vspace{-0.5in}
\caption{Luminosity and time evolution of the atmosphere of a planet with fixed core mass $M_{\rm c}=5 M_{\oplus}$ and located at $a=60$ AU. The analytic result is plotted for comparison. The luminosity is initially high, then decreases as the atmosphere grows in mass and the radiative zone becomes deeper. Growth is accelerated as a result.}
\label{fig:Ltplot}
\end{figure}

@@

We have shown in section \ref{profiles} that growth is continuously accelerated once the self-gravity of the atmosphere becomes important. This is due to both the decrease in luminosity and to the fact that less and less binding energy per unit of added mass is released as the envelope grows in mass, i.e. $-dE/dM$ becomes small. The right-hand side of equation (\ref{eq:coolingglobal}) therefore decreases with increasing atmosphere mass; in our calculations, we have found that it eventually becomes negative around the time when $M_{\rm{atm}} \sim M_{\rm{c}}$. This translates into a negative luminosity from the radiative-convective boundary. Physically, the luminosity $L$ represents the total heat flux through the surface defined by the radiative-convective boundary. The heat flux due to convective motions is always directed outwards and hence positive. The temperature gradient $dT/dr$ is negative, as the temperature decreases towards the outside, and the heat flux due to radiative diffusion is also positive as a result (see equation \ref{eq:structd}). This implies that, in the absence of planetesimal accretion and internal heat sources, there is no physical mechanism that can justify a negative luminosity. The quasi-static, constant luminosity approximation breaks down at this point and a more detailed, time-dependent model is necessary. However, we will show in section \ref{critical} that the existence of this unphysical result does not affect our estimates of the atmosphere evolution time or critical core mass, as the unphysical regime occurs after the slowest growth stage. %However, we have shown in section \ref{growthtime} that in all cases presented here, growth is slowest before $M_{\rm{atm}} \sim M_{\rm c}$, and the estimate of the time to initiate runaway gas accretion is therefore insensitive to the exact choice of crossover atmosphere mass to order of magnitude.

@@

Standard static calculations of giant planet formation that consider planetesimal accretion as the main source of luminosity (e.g., \citealt{mizuno78}, \citealt{stevenson82}, \citealt{rafikov06}) assume that the atmosphere is at all times in a steady state at which the luminosity due to planetesimal accretion is fully radiated away by the atmosphere. The core and the atmosphere grow simultaneously, and the mass of the envelope is a function of the core mass. As the atmosphere mass becomes comparable to the mass of the solid core, hydrostatic balance no longer holds and runaway gas accretion commences. Therefore, for a given set of disk conditions, there is a maximum mass the core can have before the onset of unstable accretion: this is defined as the critical core mass. %a single value for the critical core mass $M_{\rm{crit}}$ is found, when the mass of the gaseous envelope is equal to the mass of the core. This is the maximum mass the core can have before unstable gas accretion commences.

>>>>

\newcommand{\Eq}[1]{equation\,(\ref{#1})}
\newcommand{\Eqs}[2]{equations (\ref{#1}) and~(\ref{#2})}
\newcommand{\Eqss}[2]{equations (\ref{#1})--(\ref{#2})}
\newcommand{\Eqsss}[3]{equations (\ref{#1}), (\ref{#2}) and~(\ref{#3})}
\newcommand{\App}[1]{Appendix~\ref{#1}}
\newcommand{\Sec}[1]{Sect.~\ref{#1}}
\newcommand{\Chap}[1]{Chapter~\ref{#1}}
\newcommand{\Fig}[1]{Fig.~\ref{#1}}
\newcommand{\Figs}[2]{Figs.~\ref{#1} and \ref{#2}}
\newcommand{\Figss}[2]{Figs.~\ref{#1}--\ref{#2}} 
\newcommand{\Tab}[1]{Table \ref{#1}}

@@

Standard core accretion models (e.g., \citealt{stevenson82}, \citealt{rafikov06}) assume that the core and the atmosphere grow simultaneously, and that planetesimal accretion deposits enough heat to alter the luminosity of the atmosphere, increasing the core mass required for the atmosphere to collapse. These studies consider that the planet atmosphere is in a steady state at all times, where all the incoming energy from planetesimal accretion is radiated away by the envelope. The core and the atmosphere grow simultaneously, and the mass of the envelope is a function of the core mass. As the atmosphere mass becomes comparable to the mass of the solid core, hydrostatic balance no longer holds and runaway gas accretion commences. For a set of disk conditions it is therefore found that there is one well-defined core mass past which hydrostatic solutions no longer exist and unstable atmosphere collapse occurs: this is defined as the ``critical core mass''.  Standard calculations find the critical core mass to be about $10 M_{\oplus}$ (e.g., \citealt{stevenson82}, \citealt{rafikov06}).

%Standard static calculations of giant planet formation that consider planetesimal accretion as the main source of luminosity (e.g., \citealt{mizuno78}, \citealt{stevenson82}, \citealt{rafikov06}) assume that the atmosphere is at all times in a steady state at which the luminosity due to planetesimal accretion is fully radiated away by the atmosphere.  Therefore, for a given set of disk conditions, there is a maximum mass the core can have before the onset of unstable accretion: this is defined as the critical core mass.

@@

The time dependence of the atmosphere structure equations may be taken into account either explicitly or implicitly. Some previous studies of atmosphere accretion (e.g., \citealt{stevenson82}, \citealt{wuchterl93}, \citealt{rafikov06}) consider static envelopes, in which the luminosity is solely supplied by planetesimal accretion and fully radiated away by the atmosphere. In other studies, the time evolution is explicitly taken into account and full time dependent models are developed (e.g., \citealt{ikoma00}). We follow an intermediate approach and consider quasi-static evolution. Our model for the atmosphere growth time is described in section \ref{cooling}.

@@

\begin{subeqnarray}
\label{eq:struct}
\frac{d P}{d r}&=&-\frac{G m}{r^2}\rho \slabel{eq:structa} \\
\frac{d m}{d r}&=&4 \pi r^2 \rho\slabel{eq:structb} \\
\frac{d T}{d r}&=&\nabla \frac{T}{P}\frac{d P}{d r}\slabel{eq:structc} \\
\frac{d L}{d r}&=&4 \pi r^2 \rho (\epsilon + \epsilon_g)\slabel{eq:structd}, 
\end{subeqnarray}

\noindent where $r$ is the radial coordinate, $P$, $T$ and $\rho$ are the gas pressure, temperature and density, respectively, $m$ is the mass enclosed by the radius $r$, $L$ is the luminosity from the surface of radius $r$, $\epsilon$ is the rate at which heat is generated due to internal sources, and $\epsilon_g \equiv -T \frac{ds}{dt}$ represents the heating per unit mass due to gravitational contraction, with $s$ the specific gas entropy. In our study we assume that there are no internal heat sources, and therefore $\epsilon=0$. The temperature gradient $\nabla \equiv \frac{d \ln T}{d \ln P}$ has different expressions depending on the primary means of energy transport throughout the atmosphere. We assume that energy can be transported either through radiation or convection. When the luminosity is carried by radiative diffusion, the temperature gradient is given by

@@

\noindent where $\sigma$ is the Stefan-Boltzmann constant and $\kappa$ is the opacity. Equation (\ref{eq:delrad}) is valid when the atmosphere is optically thick, which is true throughout the entire atmosphere for our choice of opacity. Alternatively, when energy is transported outwards through convective motions, the temperature gradient becomes

@@

\noindent with  $\mathcal{R}$ the reduced gas constant: $\mathcal{R}=k_b/(\mu m_p)$, $k_b$ the Boltzmann constant and $m_p$ the proton mass. The gas constant $\mathcal{R}$ depends on the mean molecular weight of the gas $\mu$, which is determined by the relative abundance of hydrogen and helium in the atmosphere. In this study we assume that the helium fraction only affects the molecular weight of the gas, and not the adiabatic index.% We show in section \ref{critcore} that this assumption does not affect the qualitative behavior of our solutions.

@@

\begin{equation}
\label{eq:RHill}
R_{\rm H} \equiv a \Big(\frac{M_{\rm p}}{3 M_{\odot}}\Big)^{1/3}.
\end{equation}

\noindent Outside the Hill radius, the gravity of the planet is too weak to affect the nebular gas. Consequently, the planet can only accrete gas that lies within its Hill sphere. \\

@@

\begin{subeqnarray}
\label{eq:diskparam}
\Sigma\di&=&2200 F_{\Sigma} (a/\text{AU})^{-3/2}\,\, \text{g cm}^{-2} \slabel{eq:diska}\\
T\di &=& 120 F_T (a/\text{AU})^{-3/7} \,\,K, \slabel{eq:diskb}
\end{subeqnarray}

\noindent with $a$ the semi-major axis, and $F_{\Sigma}$ and $F_T$ normalization factors that adjust the disk mass and temperature relative to the minimum mass solar nebula (MMSN). In this study, we assume $F_{\Sigma}=F_T=1$. The resulting mid-plane pressure and disk scale height are given by

\begin{equation}
\label{eq:Pd}
P\di=110  F_{\Sigma} \sqrt{F_T m_*} (a/\text{AU})^{-45/14} \,\, \text{dyn cm}^{-2}
\end{equation}
and
\begin{equation}
H=0.022 \sqrt{F_T} (a/\text{AU})^{9/7}\,\, \text{AU},
\end{equation}

@@

\noindent where the total luminosity $L$ is balanced by the rate of heat generation $\Gamma$ (due to nuclear fusion in the case of a star) and the rate of change of the total energy of the sphere $\dot{E}$. The total energy of the sphere $E$ is given by the sum of the gravitational and internal energies: $E=E_{\rm G}+U$, with

\begin{equation}
\label{eq:Eg}
E_{\rm G}=-\int_{M\co}^M \frac{G m}{r} dm,
\end{equation}
and
\begin{equation}
\label{eq:U}
U=\int_{M\co}^M u\, dm,
\end{equation}
where $u$ is the internal energy per mass.

@@

This cooling model applies at any radius $R$ where the mass enclosed is $M$, for example the Bondi radius or the Hill radius. The $L$, $\Gamma$ and $\dot{E}$ terms are the same as in equation (\ref{eq:coolingstar}), and include the atmosphere from the core to the top boundary. In what follows we explain the additional terms in the cooling equation (\ref{eq:coolingglobal}). The first term $L\co$ is the luminosity from the solid core, and may include planetesimal accretion and radioactive decay. Since planetesimal accretion is negligible in our model, we set $L\co=0$. The fourth term on the right-hand side of equation (\ref{eq:coolingglobal}) accounts for the fact that gas is being brought in from a finite radius $R$ with a specific accretion energy $e_{\mathrm{acc}}$ given by $e_{\mathrm{acc}}=u-G M/R$. The last term represents the work done on a mass element as the atmosphere contracts. In the partial derivative the enclosed mass variable $m$ is kept fixed, and the derivative is evaluated at the surface with total enclosed mass $M$. We can think of this as rewriting the atmosphere structure equations (\ref{eq:struct}) in terms of $m$ rather than $r$, and taking into account that the thermodynamic quantities now also depend on the time $t$ (whereas time dependence is neglected in our static profiles). As such, $\partial / \partial t$ is the partial time derivative taken at constant mass.

 %As we now take into account the time-dependence of the structure equations (\ref{eq:struct}),  In the

 %In the cooling model given by equation (\ref{eq:coolingglobal}), the enclosed mass variable $m$ is kept fixed.  and take into account the time-dependence of the thermodynamic variables, which we neglected when obtaining static atmosphere models from equation (\ref{eq:struct}). This is equivalent to rewriting equation set (\ref{eq:struct}) with $m$ as the variable of integration; the partial

%In the partial derivative we are keeping fixed the enclosed mass variable $m$ and we evaluate this term at the surface with total enclosed mass $M$. This is equivalent to rewriting the structure equations (\ref{eq:struct}) with $m$ as the variable of integration, and taking into account the time dependence of the thermodynamic variables explicitly.   The partial derivative $\partial / \partial t$ is therefore taken at fixed mass.

%The partial derivative accounts for the fact that we are calculating this term at a fixed atmosphere mass, rather than total envelope mass that includes the mass accreted through the upper boundary.

The global cooling equation (\ref{eq:coolingglobal}) can be derived from the local cooling equation (\ref{eq:structd}) and the virial theorem, which relates the internal and gravitational energy of the planet as follows:
\begin{equation}
\label{eq:virial}
E_G=-3 \int_{M_c}^M \frac{P}{\rho} dm + 4 \pi (R^3 P_M-R\co^3 P\co),
\end{equation}
where $P_M$ is the pressure at radius $R$ where the mass enclosed is $M$ and $P\co$ is the pressure at the core. The virial theorem is derived from the structure equations (\ref{eq:struct}). We defer the explicit derivation of equation (\ref{eq:coolingglobal}) to \App{sec:virial}.

%The global cooling equation (\ref{eq:coolingglobal}) can be derived starting from the local cooling balance.

%Moreover, a self-gravitating protoplanetary atmosphere has to satisfy virial equilibrium. The virial theorem is derived by integrating the equation of hydrostatic balance (\ref{eq:structa}) and has the form

%\begin{equation}
%\label{eq:virial}
%E_G=-3 \int_{M_c}^M \frac{P}{\rho} dm + 4 \pi (R^3 P_M-R_c^3 Pc)
%\end{equation}
%
%Similarly to the cooling equation, it applies at any radius $R$ where the total mass enclosed is $M$. For an ideal gas of adiabatic index $\gamma$, equation (\ref{eq:virial}) leads to the following expression for the total energy:
%
%\begin{subeqnarray}
%E&=&(1-\xi)U+4 \pi (R^3 P_M-R_c^3 Pc) \slabel{eq:vira} \\
%&=&\frac{\xi-1}{\xi}E_G+\frac{4 \pi}{\xi} (R^3 P_M-R_c^3 Pc) \slabel{eq:virb}
%\end{subeqnarray}
%
%The global cooling equation (\ref{eq:coolingglobal}) can therefore be derived using equations(\ref{eq:coolinglocal}) and (\ref{eq:virial}). A detailed derivation is deferred to  \App{sec:virial}.

@@

In the next subsections we derive the structure and cooling behavior of a two-layer non self-gravitating atmosphere, deferring some of the calculations to \App{sec:analytic}. We estimate the temperature and pressure corrections at the radiative-convective boundary due to the fact that the outer radiative region of the atmosphere is not purely isothermal in section \ref{RCBcorr}. Finally, we derive the time evolution of the atmosphere and determine the minimum core mass for runaway gas accretion in section \ref{coolingan}.

@@

We use the boundary conditions and opacity laws described in section \ref{struct}. The numerical integration is performed through the shooting method (e.g., \citealt{kippenhahn90}, \citealt{press92}), which solves the boundary value problem by reducing it to an initial value problem: trial values are chosen for the parameters at one of the boundaries, then the equations are integrated and the resulting values at the other boundary are compared to the actual boundary conditions. The procedure is repeated until convergence is achieved. We start with a total atmosphere mass $M_i$, where the index $i$ labels each evolutionary stage. For this mass we guess a trial luminosity $L=L_{\mathrm{guess}}$. The assumption of constant luminosity throughout the radiative zone sets the right hand side of equation (\ref{eq:structd}) to zero, i.e. the time dependence is neglected. We integrate the structure equations (\ref{eq:struct}) inwards from $R_{\mathrm{out}}=R_{\rm H}$, with the boundary conditions $T(R_{\mathrm{out}})=T_{\rm d}$ and $P(R_{\mathrm{out}})=P_{\rm d}$. The numerical integration gives a value for the core mass implied by the trial solution $M_{\rm c}=M_{\rm{c, guess}}$. We adjust $L_{\mathrm{guess}}$ until $M_{\rm{c,guess}}$ converges to the actual core mass $M_{\rm c}$.

The time evolution of the atmosphere is obtained using the global cooling model described in section \ref{cooling}. We neglect the luminosity due to planetesimal accretion and direct heat generation; as such, the first two terms in equation (\ref{eq:coolingglobal}) are set to zero. Equation (\ref{eq:coolingglobal}) is then evaluated at the RCB, since we assume that luminosity is primarily generated in the convective zone. In what follows, $X\cb$ denotes quantity $X$ evaluated at the RCB. The time step $\Delta t_i$ between two consecutive models of masses $M_i$ and $M_{i+1}$ is calculated as follows:

\begin{eqnarray}\nonumber
\label{eq:dti}
<L>_i \Delta t_i&=&-\Delta E\cb_{,i} + <e_{\mathrm{acc, RCB}}>_i \Delta M\cb_{, i} \\
&&  - <P\cb>_i \Delta V\cb_{, M, i} 
\end{eqnarray}

@@

We consider planets embedded in a protoplanetary disk around a Sun-like star and with a temperature and pressure profile described by equations (\ref{eq:diskb}) and (\ref{eq:Pd}), respectively. We are primarily interested in the outer regions of the disk, so we generate atmosphere models at distances between 5 and 100 AU in our fiducial disk. We emphasize that the location in the disk only affects the temperature and pressure of the nebular gas. As our model applies only when planetesimal accretion occurs too slowly to substantially change the mass of the core, we consider the evolution of an atmosphere around a core of fixed mass.

@@

\begin{figure}[h]
\centering
\includegraphics[width=0.5\textwidth]{../../figs/ModelAtmospheres/RadSelfGravPoly/PaperFigs/PTm_profiles_v2.pdf}
%\vspace{-0.5in}
\caption{Example pressure, temperature and total mass (core plus atmosphere) profiles as a function of radius at $a=60$ AU and for a fixed core mass $M_{\rm c}=5 M_{\oplus}$. The circles and triangles mark the locations of the Bondi radius and of the radiative-convective boundary, respectively. The solid, dotted and dashed lines correspond to profiles with total mass (core and atmosphere) of $5.12 M_{\oplus}$, $5.85 M_{\oplus}$ and $9.22 M_{\oplus}$, respectively. The 5.12 $M_{\oplus}$ and 9.22 $M_{\oplus}$ values represent the minimum and maximum total mass, respectively, that our model allows for this particular choice of disk parameters and core mass.}
\label{fig:profiles}
\end{figure}

@@

In what follows we discuss the time evolution of the envelope. We use the global cooling model developed in section \ref{cooling} to obtain an evolutionary history from series of static atmosphere profiles, using the numerical procedure described in section \ref{twolayer}. Figure \ref{fig:Ltplot} shows the luminosity evolution and the elapsed time as a function of atmosphere mass for the same choice of parameters as in Figure \ref{fig:profiles}, $a=60$ AU and $M_{\rm c}=5 M_{\oplus}$. The radiative region of the atmosphere is initially thin, which results in a relatively small number of steps that photons need to diffuse and hence a high luminosity (see equation \ref{eq:Lcb}). The luminosity decreases as the atmosphere mass grows and the radiative region becomes deeper. For comparison, we also plot the analytic results derived in section \ref{analytic}. We find only a brief agreement between the analytic and numerical results. In the early stages, the radiative zone is too shallow and the assumption that $P_{\rm RCB} \gg P_{\rm d}$ used in the analytic model breaks down. As the mass of the atmosphere grows, the neglect of self-gravity in the analytic model results in a lower luminosity (equation \ref{eq:Lcb}) and therefore a longer cooling time when compared to the numerical solution. While our analytic model does not provide an accurate quantitative description of the atmosphere evolution, we will show in section \ref{critical} that it predicts the correct scalings for the atmosphere crossover time and critical core mass.  %We see that self-gravity starts to matter

\begin{figure}[h]
\centering
\includegraphics[width=0.5\textwidth]{../../figs/ModelAtmospheres/RadSelfGravPoly/PaperFigs/Lt_profiles_v2.pdf}
%\vspace{-0.5in}
\caption{Luminosity and time evolution of the atmosphere of a planet with fixed core mass $M_{\rm c}=5 M_{\oplus}$ and located at $a=60$ AU. The analytic result is plotted for comparison. The luminosity is initially high, then decreases as the atmosphere grows in mass and the radiative zone becomes deeper.}
\label{fig:Ltplot}
\end{figure}

@@

We have shown in section \ref{profiles} that growth is continuously accelerated once the self-gravity of the atmosphere becomes important. This is due to the fact that less and less binding energy per unit of added mass is released as the envelope grows in mass, i.e. $-dE/dM$ becomes small. The right-hand side of the cooling equation (\ref{eq:coolingglobal}) is dominated by the energy loss rate $-\dot{E}$ (as we show in \App{sec:analytic}), and as a result the right-hand side of equation (\ref{eq:coolingglobal}) decreases with increasing atmosphere mass; in our calculations, we have found that it eventually becomes negative around the time when $M_{\rm{atm}} \sim M_{\rm{c}}$. This translates into a negative luminosity from the radiative-convective boundary. Physically, the luminosity $L$ represents the total heat flux through the surface defined by the radiative-convective boundary. The heat flux due to convective motions is always directed outwards and hence positive. The temperature gradient $dT/dr$ is negative, as the temperature decreases towards the outside, and the heat flux due to radiative diffusion is also positive as a result (see equation \ref{eq:structd}). This implies that, in the absence of planetesimal accretion and internal heat sources, there is no physical mechanism that can justify a negative luminosity. The quasi-static, constant luminosity approximation breaks down at this point and a more detailed, time-dependent model is necessary. However, we will show in section \ref{critical} that the existence of this unphysical result does not affect our estimates of the atmosphere evolution time or critical core mass, as the unphysical regime occurs after the slowest growth stage. %However, we have shown in section \ref{growthtime} that in all cases presented here, growth is slowest before $M_{\rm{atm}} \sim M_{\rm c}$, and the estimate of the time to initiate runaway gas accretion is therefore insensitive to the exact choice of crossover atmosphere mass to order of magnitude.

@@

 %a single value for the critical core mass $M_{\rm{crit}}$ is found, when the mass of the gaseous envelope is equal to the mass of the core. This is the maximum mass the core can have before unstable gas accretion commences.
